---
title: "Linear models 1: introduction and one-way ANOVA"
output: 
  learnr::tutorial:
    progressive: false
    theme: cerulean
    highlight: pygments
    css: css/test.css
    code_folding: hide
runtime: shiny_prerendered
author: Rob Knell
description: >
  Analysis of variance explained: how to partition variance and why this is so useful for telling us about differences between means.
---

```{r setup, include=FALSE}r
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```


## Partitioning variance

There's a fundamental problem in statistics which is that as the number of statistical tests you do increases, so does the probability of at least one *type 1 error* --- a false positive result where you detect a significant difference which does not really exist. With an individual statistical test with the significance level set at *p = 0.05* the probability of a type 1 error is 

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

