---
title: "Linear models 1: introduction and one-way ANOVA"
output: 
  learnr::tutorial:
    progressive: false
    theme: cerulean
    highlight: pygments
    css: css/test.css
    code_folding: hide
runtime: shiny_prerendered
author: Rob Knell
description: >
  Analysis of variance explained: how to partition variance and why this is so useful for telling us about differences between means.
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```


## Partitioning variance

Let's look at some example data. These come from a paper by Minggang Wang and coauthors published in the *New Phytologist* in 2018[^1]. Wang *et al.* were studying the phenomenon of *plant-soil feedback* (PSF) whereby aspects of plant biology are affected by the presence of other plants in the soil before the one in question. The study organism in this case is ragwort, *Jacobea vulgaris*, which is an important and toxic weed in arable fields in Europe and is also highly invasive in other parts of the world.

As part of a much larger and more wide-ranging study of the effects of PSY on ragwort, Wang *et al.* grew individual ragwort plants in soil that was either sterile, inoculated with "conditioned" whole soil from a pot which a ragwort plant had previously grown in or inoculated with water which was mixed with conditioned soil and then filtered through filters of either 1000, 20, 5 or 0.5 $\mu$m mesh size. The first of these would exclude soil animals over 1mm in size, the 20 $\mu$m filter would remove most of the very small animals that live in soil such as nematodes and collembola, the 5 $\mu$m mesh would exclude most fungi and the 0.2 $\mu$m mesh would filter out most bacteria. Each pot had a single seedling grown in it for 5 weeks after which they were harvested and a variety of measurements taken including the root biomass. Let's load the data and visualise them.

```{r}
ragwort <-
  read.csv(
    "https://github.com/rjknell/Linear_models/raw/master/data/Wang_et_al_plant_only.csv",
    stringsAsFactors = TRUE
  )

str(ragwort)
```




[^1]: Wang, M., Ruan, W., Kostenko, O., Carvalho, S., Hannula, S.E., Mulder, P.P.J., Bu, F., van der Putten, W.H. & Bezemer, T.M. (2019) Removal of soil biota alters soil feedback effects on plant growth and defense chemistry. The New phytologist, 221, 1478–1491.


